---
title: "Phylogenetic Tree Inspection and Rooting (abridged)"
author: "Calla Bush St George"
date: "`r Sys.Date()`"
output:
  html_document: 
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      fig.path = "../figures/03_Phylogenetic_Tree")
```


# Setting the Environment

## Set the seed
```{r set-seed}
set.seed(238428)
```

## Load packages
```{r load-packages}
pacman::p_load(tidyverse, phyloseq, ggtree, phytools, 
               install = FALSE)
```

# Goals

1. Load the FastTree unrooted tree.
2. Add tree to phyloseq object.
3. Visualize and inspect tree with ggtree.
4. Prune ASVs, if needed.
5. Root our tree.
6. Combine new tree with a phyloseq objet.
7. Save 2 phyloseq objects: 1. Unrooted tree phyloseq object, 
                            2. Rooted tree phyloseq object.

## Load Data files
```{r load-data}
# Preprocessed phyloseq object
load("data/02_PreProcessing/raw_preprocessed_physeq.RData")
raw_preprocessed_physeq

# Load in the tree!
unrooted_tree <- read.tree("data/03_Phylogenetic_Tree/ASVs_unrooted.tree")
unrooted_tree
```

# Merge Phyloseq
```{r merge-physeq}
ntaxa(raw_preprocessed_physeq)
ntaxa(unrooted_tree)

# Merge the tree with the phyloseq object
unrooted_physeq <-
  merge_phyloseq(raw_preprocessed_physeq, unrooted_tree)

unrooted_physeq
```

```{r plot-tree-unrooted}
# Make a basic tree
kingdom_node_tree <-
  ggtree(unrooted_physeq) +
  # Color by kingdom
  geom_tippoint(mapping = aes(color = Kingdom)) +
  scale_color_manual(values = c("goldenrod1", "cornflowerblue", "gray")) +
  # Add title
  labs(title = "Unrooted Tree") + 
  # Move legened to the bottom
  theme(legend.position = "bottom") +
  # Add node labels
  geom_text(aes(label = node), hjust = 0.5, vjust = -0.3, size = 2)
kingdom_node_tree
```




# Session Information
```{r session-info}
devtools::session_info()
```